Linking assumptions and number of response options modulate inferred scalar implicature rate

### 2018-05-09 writeup.Rmd
edits to general discussion section

Because $\theta$ is a free parameter in our model, we have a straightforward understanding of between-subject response variation on a given trial of our experiment according to this linking hypothesis. **JD: Hold on, Brandon, what's the link between being a free parameter hand having an understanding of between-subject response variation?** That is, we can attribute this variation to individual differences vis a vis the value of the probability threshold against which utterances are evaluated in a given world state. Moreover, the model of a responder in the binary condition is extended intuitively to the condition where participants had three response options. In this case, we allow for two probability thresholds: $\theta_1$, the minimum standard for an utterance in a given world state to count as "right", and $\theta_2$, the minimum standard for "neither". Thus, in the ternary condition, R(u, s, $\theta_1$ , $\theta_2$) is “right” iff $S_1$(u | s) > $\theta_1$ and “neither” iff $\theta_1$ > $S_1$(u | s) > $\theta_2$.
To fully generalize the model to our five experimental conditions, we say that $R$ takes as its input an utterance $u$, a world state $s$, and a number of threshold variables dependent on a variable $c$, corresponding to the experimental condition in which the participant finds themself (e.g. the range of possible responses available to $R$).

@@

The above analysis is a proof of concept for the following idea: by relaxing the assumptions of the traditional view of scalar implicature (namely, that scalar implicatures either are or are not calculated, and that behavior on sentence verification tasks directly reflects this binary interpretation process), we can propose quantitative models of the variation in behavior we observe in experimental settings. **JD: Brandon, I feel what's still missing here is a demonstration that this is a _reasonable_ linking function. Maybe we can add the point that with this linking function, there's nothing inherently special about the implicature conditions -- the same linking function can work for any of the eight conditions in our experiment, because they're all the result of a participant having to make a behavioral choice in the face of an underlying speaker probability.**

However, this way of understanding behavior in truth value judgment tasks also raises new analytic and empirical questions which we must leave to future work. For instance, given the experimental results reported here, what is the range of plausible values for, e.g., the single probability threshold $\theta$, which separates "right" from "wrong" utterances in the binary condition? Bayesian statistical methods provide us the means for estimating the value of $\theta$ in our RSA-based model, by allowing us to recover a posterior distribution of possible values of $\theta$ given an assumption of a uniform prior and our empirically-observed data, i.e., the actual rate of response of "right" on each trial of the binary condition of the experiment. We may also ask, for example, whether the speaker probability threshold for an utterance being "right" is relatively constant across conditions or changes systematically with the introduction of more response options. Again, Bayesian data analysis allows us to explore this question by investigating the extent to which the inferred posterior distribution of, e.g., possible values of the threshold for "right" in the binary condition of the experiment differs from the inferred posterior distribution of possible values of the threshold for "right" in the ternary, quatenary, and quinary conditions.

Lastly, the linking analysis proposed here is just one in the space of possible analyses when traditional assumptions about scalar implicature are relaxed. For example, one might reject this threshold-based analysis in favor of one whereby responses are the outcomes of sampling on the (pragmatic speaker or pragmatic listener) probability distributions provided by an RSA model. We again must leave this investigation to future work, but for the time being we emphasize that such quantitative, data-driven and systematic model criticism/comparison is made available to researchers in experimental pragmatics by revising core assumptions about the nature of scalar implicature. Though we no longer have a crisp notion of scalar implicature as something that is or is not 'calculated' in interpretation, we have new flexibility to explicitly discuss categorical behavior in experimental settings.

>>>>

The model of a responder in the binary condition is extended intuitively to the condition where participants had three response options. In this case, we allow for two probability thresholds: $\theta_1$, the minimum standard for an utterance in a given world state to count as "right", and $\theta_2$, the minimum standard for "neither". Thus, in the ternary condition, R(u, s, $\theta_1$ , $\theta_2$) is “right” iff $S_1$(u | s) > $\theta_1$ and “neither” iff $\theta_1$ > $S_1$(u | s) > $\theta_2$.
To fully generalize the model to our five experimental conditions, we say that $R$ takes as its input an utterance $u$, a world state $s$, and a number of threshold variables dependent on a variable $c$, corresponding to the experimental condition in which the participant finds themself (e.g. the range of possible responses available to $R$).

@@

**The following is all new** In an RSA model, $S_1$(u | s) will be defined for any possible combination of utterance and world state. One upshot of this is that for the purposes of our linking hypothesis, participants are modeled as employing the same heuristic - does $S_1$(u | s) exceed the threshold? - in both 'implicature' and 'non-implicature' conditions of a truth value judgment task experiment. That is, participants never evaluate utterances directly on the basis of logical truth or falsity: for example, our blindfolded character Bob's guess of "cat and dog" on a cat and dog card trial is "right" to the vast majority of participants not because the guess is logically true but because $S_1$("cat and dog" | cat and dog) is exceedingly high.

For further illustration, we use our definition of a pragmatically-competent speaker $S_1$ (as defined above) to calculate the speaker probabilities of utterance for "cat", "dog", "cat and dog", and "elephant" - given either a cat on the card, or both a cat and a dog on the card. In calculating these probabilities, we assume that the space of possible utterances is the set of utterances made by Bob in our experiment (i.e. any possible single, disjunctive, or conjunctive guess involving "cat", "dog", or "elephant"). For the purposes of our model, we assume a uniform cost term on all utterances. We furthermore assume that the space of possible meanings corresponds to the set of possible card configurations that a participant may have seen in our experiment, and that the prior probability distribution over these world states is uniform. Lastly, we set $\alpha$ - the speaker rationality parameter - to 1. The resulting probability values are presented in Figure 8. **not sure that this is the right figure number**

```{r message=FALSE, warning=FALSE, fig.cap = "Speaker probabilities of utterances on the exhaustive and scalar trials, as obtained using the model described in this section."}

speaker_probs <- read.csv("models/webppl_models/speaker_probs.csv")

speaker_probs %>%
  ggplot(aes(x=u, y=prob)) +
  geom_bar(stat = "identity", position="dodge", width = 0.6) +
  facet_wrap(~s) +
  labs(x="", y= paste("Speaker probability ", "S(u | s)"))+
  theme_few() +
  theme(text = element_text(size=12), axis.text.x = element_text(angle = 45, hjust = 1))
  scale_fill_manual(values = c("red4", "springgreen3"), guide=FALSE)
```

The linking hypothesis under discussion assumes that speaker probabilities of utterance given meaning are invariant across a) our four different experimental conditions, b) across participants, and c) within participants (that is, participants are not capable of updating their $S_1$ distribution in a local discourse context). We note that the assumption (b) may conceivably be relaxed by allowing one or more of the parameters in the model - including the prior probability over world states $P_w$, the cost function on utterances $C$, or the rationality parameter $\alpha$ - to vary across participants. We also note that assumption (c) in particular is in tension with a growing body of empirical evidence that semantic and pragmatic interpretation is modulated by rapid adaptation to the linguistic and social features of one's interlocutors (c.f **TF Jaeger, Kurumada, Schuster & Degen in prep, et seq**).

However, if we should like to keep the above assumptions in place, then we must look elsewhere to explain the observed variation in our experimental data. In particular, this linking hypothesis, coupled with our assumptions, commits us to explaining variation in the data in terms of the threshold parameters of our responder model $R$. Consider first the variation in response across different experimental conditions on a given trial, e.g. evaluation of a guess of "cat and dog" when the card contains both a cat and a dog. The variation in the proportion of responses of "right" on this trial between the binary, ternary, quaternary, and quinary conditions indicates that the threshold value for "right" responses must vary across conditions; that is, we predict that the $\theta$ of the binary condition will differ from, e.g., the $\theta_1$ of the ternary condition as well as the $\theta_1$ of the quaternary condition. We also observed variation in response on this trial within a single condition (for example, a sizeable minority of participants responded "wrong" to this trial in the binary condition). Thus, this linking hypothesis is committed to the notion that threshold values may vary across participants, such that a speaker probability of utterance $S_1$(u | s) can fall below $\theta$ for some subset of participants while $S_1$(u | s) itself remains constant across participants.

Lastly, it is conceivable that for two utterances of the same conditional probability and in the same experimental condition, a participant in our experiment provided a judgment of, e.g. "right" to one utterance but "wrong" to the other. That is, it is possible that there was within-subject variation in this experiment. One way to represent such variation would be to posit that the parameterization of threshold values proceeds stochastically and that threshold values are recalibrated for every individual sentence verification task. Rather than representing a threshold as a discrete value N between 0 and 1, we can represent that threshold as a distribution over possible threshold values - with mass centered around N. Whenever an individual encounters a sentence verification task, such as a single trial of our truth value judgment task experiment, the threshold value is recalibrated by sampling from this distribution. If we allow values of $\theta$ to vary as a result of this schotastic process, for the possibility that $S_1$(u | s) sometimes falls below $\theta$ (and is otherwise above $\theta$) for a given participant.

One outstanding empirical problem is the pattern of response we observed for "cat and dog" on trials where there was only a cat on the card. Because this utterance is strictly false in this world state, it is surprising - on both the traditional view as well as on the account developed here - that participants assigned this utterance ratings above "wrong" with any systematicity. However, this is exactly what we observed, particulary in the quaternary and quinary conditions of the experiment, where a sizeable minority of participants considered this utterance "kinda right". As Figure 8 demonstrates, the conditional speaker probability of this utterance in this world state is 0; thus, there is no conceivable threshold value that would allow this utterance to ever be rated above "wrong" (on the reasonable assumption that the thresholds in our responder model $R$ should be nonzero). Any linking hypothesis will have to engage with this data point, and we leave to future work an analysis which captures participants' behavior in this condition.

For the time being, however, we present the above analysis as a proof of concept for the following idea: by relaxing the assumptions of the traditional view of scalar implicature (namely, that scalar implicatures either are or are not calculated, and that behavior on sentence verification tasks directly reflects this binary interpretation process), we can propose quantitative models of the variation in behavior we observe in experimental settings. We note that the linking analysis proposed here is just one in the space of possible analyses when traditional assumptions about scalar implicature are relaxed. For example, one might reject this threshold-based analysis in favor of one whereby responses are the outcomes of sampling on the (pragmatic speaker or pragmatic listener) probability distributions provided by an RSA model. We must leave this investigation to future work, but for the time being we emphasize that such quantitative, data-driven and systematic model criticism/comparison is made available to researchers in experimental pragmatics by revising core assumptions about the nature of scalar implicature. Though we no longer have a crisp notion of scalar implicature as something that is or is not 'calculated' in interpretation, we have new flexibility to explicitly discuss categorical behavior in experimental settings.
